---
title: "Chapter 2, Section 2.2 — A Model for Diseases with No Immunity (SIS)"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 47–51. Companion to `../../python/chapter_02/section_2_2.ipynb` —
full prose (rewrite, ELI10, key ideas, all derivations, symbol tables, intuition) is identical to
that notebook; this file reproduces the code in R.

---

## Block 2 — Phase-line analysis (Figures 2.3 and 2.4)

```{r phase-line, fig.width=12, fig.height=4.5}
f_func <- function(I, beta, gamma, N0) (beta * N0 - gamma) * I - beta * I^2

beta <- 0.001; gamma <- 0.3
rho <- gamma / beta
N0_below <- rho - 50
N0_above <- rho + 200

I_grid_a <- seq(0, N0_below * 1.3, length.out = 300)
df_a <- tibble(I = I_grid_a, f = f_func(I_grid_a, beta, gamma, N0_below))

I_grid_b <- seq(-20, N0_above * 1.1, length.out = 300)
df_b <- tibble(I = I_grid_b, f = f_func(I_grid_b, beta, gamma, N0_above))
I2_star <- N0_above - rho

p_a <- ggplot(df_a, aes(I, f)) +
  geom_line(color = "#31708f") + geom_hline(yintercept = 0) +
  annotate("point", x = 0, y = 0, color = "#3c763d", size = 3) +
  labs(title = sprintf("(a) R0=%.2f < 1 (N0=%.0f < rho=%.0f)", beta*N0_below/gamma, N0_below, rho)) +
  theme_minimal()

p_b <- ggplot(df_b, aes(I, f)) +
  geom_line(color = "#31708f") + geom_hline(yintercept = 0) +
  annotate("point", x = 0, y = 0, color = "#a94442", size = 3) +
  annotate("point", x = I2_star, y = 0, color = "#3c763d", size = 3) +
  labs(title = sprintf("(b) R0=%.2f > 1 (N0=%.0f > rho=%.0f)", beta*N0_above/gamma, N0_above, rho)) +
  theme_minimal()

gridExtra::grid.arrange(p_a, p_b, ncol = 2, top = "Figure 2.3 (recreated): phase-line analysis")
ggsave("../../chapter_02/images/sis_phase_line_R.png",
       plot = gridExtra::arrangeGrob(p_a, p_b, ncol = 2), width = 12, height = 4.5, dpi = 130)
```

```{r time-plots, fig.width=12, fig.height=4.5}
sis_rhs <- function(t, y, parms) {
  I <- y["I"]
  N0_case <- parms["N0_case"]; beta <- parms["beta"]; gamma_ <- parms["gamma_"]
  dI <- beta * I * (N0_case - I) - gamma_ * I
  list(c(dI))
}

run_case <- function(N0_case, I0_values) {
  purrr::map_dfr(I0_values, function(I0) {
    if (I0 <= 0 || I0 >= N0_case) return(NULL)
    out <- as.data.frame(ode(y = c(I = I0), times = seq(0, 200, length.out = 400),
                               func = sis_rhs, parms = c(N0_case = N0_case, beta = beta, gamma_ = gamma),
                               method = "lsoda"))
    out$I0 <- I0
    out
  })
}

res_below <- run_case(N0_below, c(5, N0_below*0.3, N0_below*0.7))
res_above <- run_case(N0_above, c(5, N0_above*0.3, N0_above*0.7))

p1 <- ggplot(res_below, aes(x = time, y = I, color = factor(I0))) +
  geom_line() + labs(title = sprintf("R0 < 1 (N0=%.0f)", N0_below), color = "I0") + theme_minimal()
p2 <- ggplot(res_above, aes(x = time, y = I, color = factor(I0))) +
  geom_line() + geom_hline(yintercept = N0_above - rho, linetype = "dashed", color = "grey40") +
  labs(title = sprintf("R0 > 1 (N0=%.0f)", N0_above), color = "I0") + theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2, top = "Figure 2.4 (recreated): monotone convergence")
ggsave("../../chapter_02/images/sis_time_plots_R.png",
       plot = gridExtra::arrangeGrob(p1, p2, ncol = 2), width = 12, height = 4.5, dpi = 130)

# Verification: monotonicity (tolerance-based, same reasoning as the Python notebook)
sol_check <- as.data.frame(ode(y = c(I = 5), times = seq(0, 200, length.out = 2000),
                                 func = sis_rhs, parms = c(N0_case = N0_above, beta = beta, gamma_ = gamma),
                                 method = "lsoda"))
diffs <- diff(sol_check$I)
tol <- 1e-6
sig <- diffs[abs(diffs) > tol]
sign_changes <- sum(diff(sign(sig)) != 0)
cat(sprintf("Sign changes among significant steps: %d (should be 0 -- monotone)\n", sign_changes))
```

## Block 3 — SIR vs. SIS comparison (does S(t) cross rho?)

```{r sir-vs-sis, fig.width=12, fig.height=4.5}
beta2 <- 0.0005; gamma2 <- 0.2
rho2 <- gamma2 / beta2
N0c <- 700; I0c <- 5; S0c <- N0c - I0c

sir_rhs <- function(t, y, parms) {
  S <- y["S"]; I <- y["I"]
  dS <- -parms["beta"] * I * S
  dI <- parms["beta"] * I * S - parms["gamma_"] * I
  list(c(dS, dI))
}
sis_rhs2 <- function(t, y, parms) {
  I <- y["I"]
  dI <- parms["beta"] * I * (parms["N0"] - I) - parms["gamma_"] * I
  list(c(dI))
}

t_eval <- seq(0, 200, length.out = 800)
sol_sir <- as.data.frame(ode(y = c(S = S0c, I = I0c), times = t_eval, func = sir_rhs,
                               parms = c(beta = beta2, gamma_ = gamma2), method = "lsoda"))
sol_sis <- as.data.frame(ode(y = c(I = I0c), times = t_eval, func = sis_rhs2,
                               parms = c(beta = beta2, gamma_ = gamma2, N0 = N0c), method = "lsoda"))
sol_sis$S <- N0c - sol_sis$I

p_sir <- ggplot(sol_sir, aes(x = time)) +
  geom_line(aes(y = S, color = "S(t)")) + geom_line(aes(y = I, color = "I(t)")) +
  geom_hline(yintercept = rho2, linetype = "dashed", color = "grey40") +
  labs(title = "SIR: S(t) CROSSES rho -> epidemic peak", color = "") + theme_minimal()
p_sis <- ggplot(sol_sis, aes(x = time)) +
  geom_line(aes(y = S, color = "S(t)")) + geom_line(aes(y = I, color = "I(t)")) +
  geom_hline(yintercept = rho2, linetype = "dashed", color = "grey40") +
  labs(title = "SIS: S(t) stays ABOVE rho -> no peak", color = "") + theme_minimal()

gridExtra::grid.arrange(p_sir, p_sis, ncol = 2)
ggsave("../../chapter_02/images/sir_vs_sis_comparison_R.png",
       plot = gridExtra::arrangeGrob(p_sir, p_sis, ncol = 2), width = 12, height = 4.5, dpi = 130)

cat(sprintf("SIR: min S(t) = %.2f (crosses rho=%.0f? %s)\n", min(sol_sir$S), rho2, min(sol_sir$S) < rho2))
cat(sprintf("SIS: min S(t) = %.2f (crosses rho=%.0f? %s)\n", min(sol_sis$S), rho2, min(sol_sis$S) < rho2))
```

## Block 4 — Bifurcation diagram

```{r bifurcation, fig.width=8, fig.height=5.5}
N0_grid <- seq(0, 2 * rho, length.out = 400)
I2_star_grid <- N0_grid - rho

bif_df <- tibble(
  N0 = N0_grid,
  I1 = 0,
  stable1 = N0_grid < rho,
  I2 = I2_star_grid
) %>% filter(I2 >= 0)

branch1_stable <- tibble(N0 = N0_grid[N0_grid < rho], I = 0)
branch1_unstable <- tibble(N0 = N0_grid[N0_grid >= rho], I = 0)

ggplot() +
  geom_line(data = branch1_stable, aes(N0, I), color = "#3c763d", linewidth = 1.2) +
  geom_line(data = branch1_unstable, aes(N0, I), color = "#a94442", linewidth = 1, linetype = "dashed") +
  geom_line(data = bif_df, aes(N0, I2), color = "#3c763d", linewidth = 1.2) +
  geom_vline(xintercept = rho, linetype = "dotted", color = "grey40") +
  labs(title = "Transcritical bifurcation: equilibria of the SIS model vs. N0",
       x = "N0 (bifurcation parameter)", y = "equilibrium I*") +
  theme_minimal()

ggsave("../../chapter_02/images/sis_bifurcation_diagram_R.png", width = 8, height = 5.5, dpi = 130)
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_02/section_2_2.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_02/exercises/section_2_2_exercises.md`; solutions:
`../../chapter_02/solutions/section_2_2_solutions.md`.*
