---
title: "Chapter 3, Sections 3.1-3.2 — Stability of Equilibria & Linearization"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 79–82. Companion to
`../../python/chapter_03/section_3_1_3_2.ipynb` — full prose (rewrite, ELI10, key ideas,
derivations, symbol tables, intuition) is identical to that notebook; this file reproduces the
code in R.

---

## Block 1 — Stable vs. asymptotically stable: undamped vs. damped oscillator

```{r stability-comparison, fig.width=11, fig.height=5.5}
undamped <- function(t, y, parms) {
  x <- y["x"]; v <- y["v"]
  list(c(v, -x))
}
damped <- function(t, y, parms) {
  x <- y["x"]; v <- y["v"]
  c <- parms["c"]
  list(c(v, -x - c*v))
}

t_eval <- seq(0, 40, length.out = 2000)
starts <- list(c(1,0), c(0.6,0), c(0.3,0))

undamped_trajs <- purrr::map_dfr(seq_along(starts), function(i) {
  st <- starts[[i]]
  out <- as.data.frame(ode(y = c(x=st[1], v=st[2]), times = t_eval, func = undamped, parms = c()))
  out$traj <- i; out
})
damped_trajs <- purrr::map_dfr(seq_along(starts), function(i) {
  st <- starts[[i]]
  out <- as.data.frame(ode(y = c(x=st[1], v=st[2]), times = t_eval, func = damped, parms = c(c=0.3)))
  out$traj <- i; out
})

p1 <- ggplot(undamped_trajs, aes(x, v, group = traj)) + geom_path(alpha=0.8) +
  annotate("point", x=0, y=0, color="red", size=2) + coord_equal() +
  labs(title = "STABLE but NOT asymptotically stable\n(undamped: orbits stay close, never converge)") +
  theme_minimal()
p2 <- ggplot(damped_trajs, aes(x, v, group = traj)) + geom_path(alpha=0.8) +
  annotate("point", x=0, y=0, color="red", size=2) +
  labs(title = "ASYMPTOTICALLY STABLE\n(damped: orbits spiral IN to the equilibrium)") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
ggsave("../../chapter_03/images/stable_vs_asymptotically_stable_R.png",
       plot = gridExtra::arrangeGrob(p1, p2, ncol = 2), width = 11, height = 5.5, dpi = 130)

sol1 <- as.data.frame(ode(y = c(x=1,v=0), times = seq(0,100,length.out=2000), func = undamped, parms = c()))
dist1 <- sqrt(sol1$x^2 + sol1$v^2)
cat(sprintf("Undamped: distance from origin ranges over [%.6f, %.6f]\n", min(dist1), max(dist1)))

sol2 <- as.data.frame(ode(y = c(x=1,v=0), times = seq(0,100,length.out=2000), func = damped, parms = c(c=0.3)))
dist2 <- sqrt(sol2$x^2 + sol2$v^2)
cat(sprintf("Damped: distance from origin at t=0: %.4f, at t=100: %.6f\n", dist2[1], tail(dist2,1)))
```

## Block 2 — Routh-Hurwitz criteria vs. direct eigenvalue computation

```{r routh-hurwitz-check}
set.seed(42)

principal_minors_sum <- function(A) {
  M1 <- A[2,2]*A[3,3] - A[2,3]*A[3,2]
  M2 <- A[1,1]*A[3,3] - A[1,3]*A[3,1]
  M3 <- A[1,1]*A[2,2] - A[1,2]*A[2,1]
  M1 + M2 + M3
}

routh_hurwitz_3x3 <- function(A) {
  tr <- sum(diag(A)); det_A <- det(A); a2 <- principal_minors_sum(A)
  (tr < 0) && (det_A < 0) && (tr*a2 - det_A < 0)
}

all_eigs_negative_real <- function(A) all(Re(eigen(A, only.values = TRUE)$values) < 0)

n_tests <- 5000
agreements <- 0
for (i in 1:n_tests) {
  A <- matrix(runif(9, -3, 3), nrow = 3)
  if (routh_hurwitz_3x3(A) == all_eigs_negative_real(A)) agreements <- agreements + 1
}
cat(sprintf("Routh-Hurwitz (3x3) matched direct eigenvalues in %d/%d trials (%.2f%%)\n",
            agreements, n_tests, 100*agreements/n_tests))

routh_hurwitz_2x2 <- function(A) (sum(diag(A)) < 0) && (det(A) > 0)
agreements_2x2 <- 0
for (i in 1:n_tests) {
  A <- matrix(runif(4, -3, 3), nrow = 2)
  if (routh_hurwitz_2x2(A) == all_eigs_negative_real(A)) agreements_2x2 <- agreements_2x2 + 1
}
cat(sprintf("Routh-Hurwitz (2x2) matched direct eigenvalues in %d/%d trials\n", agreements_2x2, n_tests))

# Callback to Sec 2.3's demography model
jacobian_Pstar_sec23 <- function(b, beta, gamma_) {
  S_star <- (b + gamma_) / beta
  I_star <- b * (beta - (b + gamma_)) / (beta * (b + gamma_))
  matrix(c(-beta*I_star - b, beta*I_star, -beta*S_star, 0), nrow = 2)
}
b <- 0.02; beta <- 0.3; gamma_ <- 0.2
J <- jacobian_Pstar_sec23(b, beta, gamma_)
cat("\nSec. 2.3 callback J(P*):\n"); print(J)
cat(sprintf("tr(J)=%.5f, det(J)=%.5f, RH predicts stable: %s\n",
            sum(diag(J)), det(J), routh_hurwitz_2x2(J)))
cat("Direct eigenvalues:\n"); print(eigen(J, only.values = TRUE)$values)
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_03/section_3_1_3_2.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_03/exercises/section_3_1_3_2_exercises.md`; solutions:
`../../chapter_03/solutions/section_3_1_3_2_solutions.md`.*
