---
title: "Chapter 3, Sections 3.7-3.8 — Uniform Persistence & Metzler Matrices/Monotone Systems"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 95–101. Companion to
`../../python/chapter_03/section_3_7_3_8.ipynb` — full prose (rewrite, ELI10, key ideas,
derivations, symbol tables, intuition) is identical to that notebook; this file reproduces the
code in R.

---

## Block 1 (§3.7) — Uniform persistence for Sec. 2.3's model

```{r persistence}
rhs <- function(t, y, parms) {
  S <- y["S"]; I <- y["I"]
  b <- parms["b"]; beta <- parms["beta"]; gamma_ <- parms["gamma_"]
  list(c(b - beta*I*S - b*S, beta*I*S - gamma_*I - b*I))
}

b <- 0.02; gamma_ <- 0.2
beta_above <- 0.5
R0 <- beta_above / (b + gamma_)
cat(sprintf("R0 = %.3f (> 1, testing uniform persistence)\n", R0))

starts <- list(c(0.999,0.0005), c(0.001,0.001), c(0.5,0.001), c(0.001,0.5), c(0.5,0.499), c(0.9,0.09))
t_tail <- seq(400, 600, length.out = 500)

min_S <- Inf; min_I <- Inf; min_gap <- Inf
for (st in starts) {
  out <- as.data.frame(ode(y = c(S=st[1], I=st[2]), times = t_tail, func = rhs,
                             parms = c(b=b, beta=beta_above, gamma_=gamma_), method = "lsoda"))
  gap <- 1 - out$S - out$I
  min_S <- min(min_S, min(out$S)); min_I <- min(min_I, min(out$I)); min_gap <- min(min_gap, min(gap))
  cat(sprintf("  start=(%.3f,%.4f): tail S in [%.4f,%.4f], tail I in [%.4f,%.4f]\n",
              st[1], st[2], min(out$S), max(out$S), min(out$I), max(out$I)))
}
cat(sprintf("\nAcross all trajectories: min S=%.4f, min I=%.4f, min(1-S-I)=%.4f\n", min_S, min_I, min_gap))
cat("All strictly positive -> consistent with UNIFORM persistence\n")

# Contrast: R0 < 1
beta_below <- 0.15
R0_below <- beta_below / (b + gamma_)
cat(sprintf("\nR0 = %.3f (< 1, persistence should FAIL)\n", R0_below))
out_below <- as.data.frame(ode(y = c(S=0.5, I=0.3), times = t_tail, func = rhs,
                                 parms = c(b=b, beta=beta_below, gamma_=gamma_), method = "lsoda"))
cat(sprintf("Tail I(t) range when R0<1: [%.6f, %.6f]\n", min(out_below$I), max(out_below$I)))
cat("I(t) -> 0 -> persistence correctly FAILS\n")
```

## Block 2 (§3.8) — Perron-Frobenius for Metzler matrices, and Theorem 3.8.3

```{r metzler-checks}
set.seed(1)

random_metzler <- function(n, diag_range = c(-5,-0.1), offdiag_range = c(0,3)) {
  A <- matrix(runif(n*n, offdiag_range[1], offdiag_range[2]), nrow = n)
  diag(A) <- runif(n, diag_range[1], diag_range[2])
  A
}

violations <- 0
n_tests <- 500
for (i in 1:n_tests) {
  n <- sample(2:5, 1)
  A <- random_metzler(n)
  eig <- eigen(A)
  s_A <- max(Re(eig$values))
  idx <- which.min(abs(eig$values - s_A))
  is_eigenvalue <- abs(Im(eig$values[idx])) < 1e-8
  v <- Re(eig$vectors[, idx])
  nonneg <- all(v >= -1e-8) || all(v <= 1e-8)
  if (!(is_eigenvalue && nonneg)) violations <- violations + 1
}
cat(sprintf("Perron-Frobenius (Metzler) check: %d violations out of %d random matrices\n", violations, n_tests))

# Theorem 3.8.3 equivalence (1) <=> (2)
agreements <- 0
for (i in 1:n_tests) {
  n <- sample(2:4, 1)
  A <- random_metzler(n, diag_range = c(-8,2))
  s_A <- max(Re(eigen(A, only.values = TRUE)$values))
  is_stable <- s_A < 0
  inv_ok <- tryCatch({
    minus_Ainv <- -solve(A)
    all(minus_Ainv >= -1e-8)
  }, error = function(e) FALSE)
  if (is_stable == inv_ok) agreements <- agreements + 1
}
cat(sprintf("Theorem 3.8.3 equivalence (1)<=>(2): %d/%d random Metzler matrices agree\n", agreements, n_tests))
```

## Block 2 continued — Theorem 3.8.5 applied to Sec. 2.5's Ross-MacDonald model

```{r theorem-3-8-5}
ross_macdonald <- function(t, state, parms) {
  x <- state["x"]; y <- state["y"]
  a <- parms["a"]; m <- parms["m"]; b1 <- parms["b1"]; b2 <- parms["b2"]
  gamma1 <- parms["gamma1"]; gamma2 <- parms["gamma2"]
  dx <- a*m*b1*y*(1-x) - gamma1*x
  dy <- a*b2*x*(1-y) - gamma2*y
  list(c(dx, dy))
}

a_param <- 0.3; b1 <- 0.5; b2 <- 0.3; gamma1 <- 0.1; gamma2 <- 0.05

for (case in list(list(m=0.3, label="R0 < 1 (all solutions -> 0)"),
                   list(m=6.0, label="R0 > 1 (globally stable p>0)"))) {
  m_param <- case$m
  R0 <- a_param^2 * m_param * b1 * b2 / (gamma1 * gamma2)
  t_end <- if (R0 < 1) 3000 else 500
  results <- purrr::map(list(c(0.1,0.1), c(0.9,0.05), c(0.05,0.9), c(0.5,0.5)), function(st) {
    out <- ode(y = c(x=st[1], y=st[2]), times = c(0, t_end), func = ross_macdonald,
               parms = c(a=a_param, m=m_param, b1=b1, b2=b2, gamma1=gamma1, gamma2=gamma2), method = "lsoda")
    out[nrow(out), c("x","y")]
  })
  cat(sprintf("%s: R0=%.2f\n", case$label, R0))
  for (r in results) cat(sprintf("  (%.6f, %.6f)\n", r[1], r[2]))
}
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_03/section_3_7_3_8.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_03/exercises/section_3_7_3_8_exercises.md`; solutions:
`../../chapter_03/solutions/section_3_7_3_8_solutions.md`.*

---

# Chapter 3 Complete

All eight sections of Chapter 3 are now built, verified numerically, and cross-checked against
their concrete applications throughout Chapter 2.
