---
title: "Chapter 3, Sections 3.5-3.6 — Phase-Line & Phase-Plane Analysis"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 87–94. Companion to
`../../python/chapter_03/section_3_5_3_6.ipynb` — full prose (rewrite, ELI10, key ideas,
derivations, symbol tables, intuition) is identical to that notebook; this file reproduces the
code in R.

---

## Block 1 (§3.5) — Logistic equation phase-line analysis

```{r logistic, fig.width=12, fig.height=4.5}
r <- 0.3; K <- 100
f_logistic <- function(N) r * N * (1 - N / K)

N_grid <- seq(-10, 130, length.out = 300)
p1 <- ggplot(tibble(N = N_grid, f = f_logistic(N_grid)), aes(N, f)) +
  geom_line(color = "#31708f") + geom_hline(yintercept = 0) +
  annotate("point", x = 0, y = 0, color = "#a94442", size = 3) +
  annotate("point", x = K, y = 0, color = "#3c763d", size = 3) +
  labs(title = "Phase-line analysis: logistic equation") + theme_minimal()

logistic_rhs <- function(t, y, parms) list(f_logistic(y["N"]))
traj_df <- purrr::map_dfr(c(1, 20, 50, 80, 120, 150), function(N0) {
  out <- as.data.frame(ode(y = c(N=N0), times = seq(0, 40, length.out = 300),
                             func = logistic_rhs, parms = c(), method = "lsoda"))
  out$N0 <- N0
  out
})
p2 <- ggplot(traj_df, aes(time, N, color = factor(N0))) + geom_line() +
  geom_hline(yintercept = K, linetype = "dashed", color = "grey40") +
  labs(title = "All solutions -> K", color = "N0") + theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
ggsave("../../chapter_03/images/logistic_phase_line_R.png",
       plot = gridExtra::arrangeGrob(p1, p2, ncol = 2), width = 12, height = 4.5, dpi = 130)
```

## Block 2 — Van der Pol oscillator: Poincare-Bendixson in action

```{r van-der-pol, fig.width=7, fig.height=6.5}
mu <- 1.0
van_der_pol <- function(t, y, parms) {
  x <- y["x"]; v <- y["v"]
  list(c(v, mu*(1 - x^2)*v - x))
}

J0 <- matrix(c(0, -1, 1, mu), nrow = 2)
cat("Jacobian at origin:\n"); print(J0)
cat(sprintf("trace=%.2f (>0 -> unstable), det=%.2f\n", sum(diag(J0)), det(J0)))
cat("Eigenvalues:\n"); print(eigen(J0, only.values = TRUE)$values)

starts <- list(c(0.01,0), c(4,0), c(0,3), c(-3,-3))
vdp_traj <- purrr::map_dfr(seq_along(starts), function(i) {
  st <- starts[[i]]
  out <- as.data.frame(ode(y = c(x=st[1], v=st[2]), times = seq(0, 40, length.out = 4000),
                             func = van_der_pol, parms = c(), method = "lsoda"))
  out$traj <- i
  out
})

p3 <- ggplot(vdp_traj, aes(x, v, group = traj)) + geom_path(alpha = 0.75) +
  annotate("point", x = 0, y = 0, color = "red", shape = 4, size = 4) +
  labs(title = "Van der Pol (mu=1): Poincare-Bendixson guarantees a periodic orbit") +
  theme_minimal()
print(p3)
ggsave("../../chapter_03/images/van_der_pol_limit_cycle_R.png", plot = p3, width = 7, height = 6.5, dpi = 130)
```

## Block 3 — Liouville's formula cross-check against Sec. 3.4's Floquet multiplier

```{r liouville}
# R lacks as convenient symbolic integration as sympy for this derivation;
# we confirm the SAME claim numerically instead -- integrate div(f) along the
# orbit p(t)=(cos t, sin t) over one period using numerical quadrature.
div_f <- function(x, y) 2 - 4*x^2 - 4*y^2   # div f(x,y) for the r'=r(1-r^2) system
div_along_orbit <- function(t) div_f(cos(t), sin(t))

integral_val <- integrate(div_along_orbit, lower = 0, upper = 2*pi)$value
cat(sprintf("Integral of div f(p(t)) over one period: %.6f (expect -4*pi = %.6f)\n",
            integral_val, -4*pi))

lambda2_predicted <- exp(integral_val)
cat(sprintf("Liouville's formula predicts lambda_2 = %.6e\n", lambda2_predicted))
cat(sprintf("Sec. 3.4's directly-computed second Floquet multiplier was: 3.48734e-06\n"))
```

## Block 4 — Bendixson's Negative Criterion: van der Pol vs. damped linear oscillator

```{r bendixson}
# div f for van der Pol: mu*(1-x^2) -- sign-changing, does NOT rule out periodic orbits
cat("div f (van der Pol) = mu*(1-x^2) -- positive for |x|<1, negative for |x|>1: NOT sign-definite\n")
cat("-> Bendixson's criterion does not rule out a periodic orbit (consistent with Block 2)\n\n")

# div f for the damped linear oscillator: constant -c -- sign-definite
c_val <- 0.3
cat(sprintf("div f (damped linear oscillator) = -c = %.2f -- sign-definite (negative) everywhere\n", -c_val))
cat("-> Bendixson's criterion correctly RULES OUT any periodic orbit (consistent with Sec. 3.1)\n")
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_03/section_3_5_3_6.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_03/exercises/section_3_5_3_6_exercises.md`; solutions:
`../../chapter_03/solutions/section_3_5_3_6_solutions.md`.*
