---
title: "Chapter 4, Section 4.1 — Curve-Fitting and the Linear Least-Squares Problem"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
```

**Source:** Li, M.Y. (2018), pp. 104–111. Companion to `../../python/chapter_04/section_4_1.ipynb`
— full prose (rewrite, ELI10, key ideas, derivations, symbol tables, intuition) is identical to
that notebook; this file reproduces the code in R.

---

## Example 1 — Linear fit

```{r example1}
x_data <- c(2, 5, 7, 8)
y_data <- c(1, 2, 3, 3)

A <- cbind(1, x_data)
b <- y_data

# Method 1: normal equations
x_hat_normal_eq <- solve(t(A) %*% A, t(A) %*% b)
cat("Method 1 (normal equations):", x_hat_normal_eq, "\n")

# Method 2: R's built-in lm()
fit <- lm(y_data ~ x_data)
cat("Method 2 (lm()):", coef(fit), "\n")

cat(sprintf("\nBook's stated answer: a0 = 2/7 = %.6f, a1 = 5/14 = %.6f\n", 2/7, 5/14))

# Orthogonality check
residual_vec <- b - A %*% x_hat_normal_eq
for (j in 1:ncol(A)) {
  dot <- sum(residual_vec * A[, j])
  cat(sprintf("(b - A*x_hat) . (column %d of A) = %.2e (should be ~0)\n", j, dot))
}
```

## Example 2 — Quadratic fit

```{r example2}
x_data2 <- c(2, -1, 6, 4)
y_data2 <- c(1, 5, 2, -1)

A2 <- cbind(1, x_data2, x_data2^2)
b2 <- y_data2

x_hat2 <- solve(t(A2) %*% A2, t(A2) %*% b2)
cat("Least-squares quadratic coefficients (a0, a1, a2):", x_hat2, "\n")
cat("Book's stated answer: a0=2.9719, a1=-1.909, a2=0.2826\n")

# Cross-check with R's poly() / lm()
fit2 <- lm(y_data2 ~ x_data2 + I(x_data2^2))
cat("lm() cross-check:", coef(fit2), "\n")
```

## Example 3 — Exponential fit via log-linearization

```{r example3}
set.seed(3)
a0_true <- 2.0; a1_true <- 0.35
t_data <- c(0.5, 1.0, 1.5, 2.0, 2.5, 3.0, 3.5)
x_true <- a0_true * exp(a1_true * t_data)
x_data3 <- x_true * exp(rnorm(length(t_data), sd = 0.05))

y_data3 <- log(x_data3)
A3 <- cbind(1, t_data)
coefs3 <- solve(t(A3) %*% A3, t(A3) %*% y_data3)
b0_hat <- coefs3[1]; a1_hat <- coefs3[2]
a0_hat <- exp(b0_hat)

cat(sprintf("True parameters:      a0=%.2f, a1=%.2f\n", a0_true, a1_true))
cat(sprintf("Recovered parameters: a0=%.4f, a1=%.4f\n", a0_hat, a1_hat))
```

## Visualizing all three fits

```{r plot-all, fig.width=15, fig.height=4.5}
p1 <- ggplot() +
  geom_point(data = tibble(x=x_data, y=y_data), aes(x, y)) +
  geom_abline(intercept = x_hat_normal_eq[1], slope = x_hat_normal_eq[2], color = "#31708f") +
  labs(title = "Example 1: linear fit") + theme_minimal()

x_curve <- seq(-2, 7, length.out = 200)
p2 <- ggplot() +
  geom_point(data = tibble(x=x_data2, y=y_data2), aes(x, y)) +
  geom_line(data = tibble(x=x_curve, y=x_hat2[1]+x_hat2[2]*x_curve+x_hat2[3]*x_curve^2),
            aes(x, y), color = "#a94442") +
  labs(title = "Example 2: quadratic fit") + theme_minimal()

t_curve <- seq(0, 4, length.out = 100)
p3 <- ggplot() +
  geom_point(data = tibble(t=t_data, x=x_data3), aes(t, x)) +
  geom_line(data = tibble(t=t_curve, x=a0_hat*exp(a1_hat*t_curve)), aes(t, x), color = "#3c763d") +
  geom_line(data = tibble(t=t_curve, x=a0_true*exp(a1_true*t_curve)), aes(t, x),
            color = "grey40", linetype = "dashed") +
  labs(title = "Example 3: exponential fit (log-linearized)") + theme_minimal()

gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
ggsave("../../chapter_04/images/least_squares_examples_R.png",
       plot = gridExtra::arrangeGrob(p1, p2, p3, ncol = 3), width = 15, height = 4.5, dpi = 130)
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_04/section_4_1.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_04/exercises/section_4_1_exercises.md`; solutions:
`../../chapter_04/solutions/section_4_1_solutions.md`.*
