---
title: "Chapter 4, Section 4.2 — Nonlinear Least-Squares Problem"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(minpack.lm)
```

**Source:** Li, M.Y. (2018), pp. 111–117. Companion to `../../python/chapter_04/section_4_2.ipynb`
— full prose (rewrite, ELI10, key ideas, derivations, symbol tables, intuition) is identical to
that notebook; this file reproduces the code in R. Note: this section additionally uses
`minpack.lm` (Levenberg-Marquardt nonlinear least squares) as the professional cross-check,
analogous to `scipy.optimize.curve_fit` in the Python notebook.

---

## Gauss-Newton implemented from scratch

```{r gauss-newton}
t_data <- c(1, 2, 3, 4, 5)
x_data <- c(4.6, 8.82, 16, 31.3, 58.5)

f_model <- function(t, a0, a1) a0 * exp(a1 * t)

jacobian_gn <- function(t, a0, a1) {
  col0 <- exp(a1 * t)
  col1 <- a0 * t * exp(a1 * t)
  cbind(col0, col1)
}

gauss_newton <- function(t, y, f, jac, theta0, n_iter = 6) {
  theta <- theta0
  history <- list(theta)
  for (k in 1:n_iter) {
    residuals <- y - f(t, theta[1], theta[2])
    J <- jac(t, theta[1], theta[2])
    delta_theta <- solve(t(J) %*% J, t(J) %*% residuals)
    theta <- theta + as.vector(delta_theta)
    history[[k+1]] <- theta
    cat(sprintf("iteration %d: theta = (%.6f, %.6f)\n", k, theta[1], theta[2]))
  }
  list(theta = theta, history = history)
}

result <- gauss_newton(t_data, x_data, f_model, jacobian_gn, theta0 = c(1, 1), n_iter = 6)
theta_final <- result$theta
cat(sprintf("\nFinal Gauss-Newton estimate: a0=%.4f, a1=%.4f\n", theta_final[1], theta_final[2]))
cat("Book's stated Gauss-Newton answer: a0~=2.431, a1~=0.636\n")
```

## Cross-checks

```{r cross-checks}
# Log-linearization (Sec. 4.1 style)
A <- cbind(1, t_data)
coefs <- solve(t(A) %*% A, t(A) %*% log(x_data))
a0_log <- exp(coefs[1]); a1_log <- coefs[2]
cat(sprintf("Solution 1 (log-linearization): a0=%.4f, a1=%.4f\n", a0_log, a1_log))
cat(sprintf("Solution 2 (Gauss-Newton, direct): a0=%.4f, a1=%.4f\n", theta_final[1], theta_final[2]))

# minpack.lm (Levenberg-Marquardt) cross-check
fit_lm <- nlsLM(x_data ~ a0 * exp(a1 * t_data), start = list(a0 = 1, a1 = 1))
cat("\nminpack.lm::nlsLM cross-check:\n")
print(coef(fit_lm))

sse <- function(a0, a1) sum((x_data - f_model(t_data, a0, a1))^2)
cat(sprintf("\nSSE (original space) at log-linearized solution: %.4f\n", sse(a0_log, a1_log)))
cat(sprintf("SSE (original space) at Gauss-Newton solution:    %.4f\n", sse(theta_final[1], theta_final[2])))
```

## Visualizing both fits

```{r plot-fits, fig.width=8, fig.height=5.5}
t_curve <- seq(0.5, 5.5, length.out = 100)
plot_df <- bind_rows(
  tibble(t = t_curve, x = f_model(t_curve, theta_final[1], theta_final[2]), curve = "Gauss-Newton"),
  tibble(t = t_curve, x = f_model(t_curve, a0_log, a1_log), curve = "log-linearized")
)

ggplot() +
  geom_point(data = tibble(t=t_data, x=x_data), aes(t, x), size = 2) +
  geom_line(data = plot_df, aes(t, x, color = curve, linetype = curve)) +
  labs(title = "Figure 4.4 (recreated): the two curves are nearly indistinguishable") +
  theme_minimal()

ggsave("../../chapter_04/images/gauss_newton_example4_R.png", width = 8, height = 5.5, dpi = 130)
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_04/section_4_2.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_04/exercises/section_4_2_exercises.md`; solutions:
`../../chapter_04/solutions/section_4_2_solutions.md`.*
