---
title: "Chapter 5, Section 5.2 — In-host Models and Backward Bifurcation"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 136–146. Companion to `../../python/chapter_05/section_5_2.ipynb`
— full prose (rewrite, ELI10, key ideas, derivations, symbol tables, intuition) is identical to
that notebook; this file reproduces the code in R. As in the Python notebook, every quantitative
claim here is verified numerically (root-finding, eigenvalues, simulation) rather than relying on
a hand-copied closed-form formula for the bifurcation thresholds.

---

## Block 1 — Feasible region

```{r feasible-region}
lam <- 8.938; K <- 1405.3
nu1 <- 0.285; nu2 <- 0.866
mu1 <- 0.126; mu2 <- 0.484
beta <- 0.000862

nu_max <- max(nu1, nu2); mu_min <- min(mu1, mu2)
coeffs <- c(-nu_max/K, (nu_max-mu_min), lam)
roots_N <- polyroot(rev(coeffs))
N_bar <- max(Re(roots_N)[abs(Im(roots_N)) < 1e-9 & Re(roots_N) > 0])
cat(sprintf("Feasible region bound N_bar = %.3f\n", N_bar))

inhost_model <- function(t, state, parms) {
  x <- max(state["x"], 0); y <- max(state["y"], 0)
  sigma <- parms["sigma"]
  dx <- lam + nu1*x*(1-(x+y)/K) - mu1*x - beta*x*y
  dy <- sigma*beta*x*y + nu2*y*(1-(x+y)/K) - mu2*y
  list(c(dx, dy))
}

for (pt in list(c(10,5), c(500,300), c(50,900), c(1,1))) {
  if (sum(pt) > N_bar) next
  out <- ode(y = c(x=pt[1], y=pt[2]), times = seq(0,500,length.out=500),
             func = inhost_model, parms = c(sigma=0.5), method = "lsoda")
  max_N <- max(out[,"x"] + out[,"y"])
  cat(sprintf("start=(%g,%g): max(x+y) = %.3f (bound N_bar=%.3f)\n", pt[1], pt[2], max_N, N_bar))
}
```

## Block 2 — Locating the bifurcation numerically

```{r bifurcation-count, fig.width=8, fig.height=4.5}
g_of_x <- function(x, sigma) (lam + (nu1/nu2)*x*(mu2 - sigma*beta*x) - mu1*x) / (beta*x)
h_of_x <- function(x, sigma) K*(1 - (mu2 - sigma*beta*x)/nu2) - x

chronic_equilibria <- function(sigma, n = 6000) {
  xs <- seq(1, K, length.out = n)
  diffs <- g_of_x(xs, sigma) - h_of_x(xs, sigma)
  roots <- list()
  for (i in 1:(length(xs)-1)) {
    if (is.finite(diffs[i]) && is.finite(diffs[i+1]) && sign(diffs[i]) != sign(diffs[i+1])) {
      r <- uniroot(function(xx) g_of_x(xx, sigma) - h_of_x(xx, sigma), c(xs[i], xs[i+1]))$root
      yv <- h_of_x(r, sigma)
      if (yv > 0.01) roots[[length(roots)+1]] <- c(r, yv)
    }
  }
  roots
}

sigma_grid <- seq(0, 0.4, length.out = 60)
counts <- sapply(sigma_grid, function(s) length(chronic_equilibria(s)))

ggplot(tibble(sigma=sigma_grid, count=counts), aes(sigma, count)) +
  geom_point(color = "#31708f") + geom_line(color = "#31708f") +
  scale_y_continuous(breaks = c(0,1,2)) +
  labs(title = "Chronic-equilibrium count vs. sigma (located numerically)", x = TeX("$\\sigma$")) +
  theme_minimal()

ggsave("../../chapter_05/images/backward_bifurcation_count_R.png", width = 8, height = 4.5, dpi = 130)

# Bisect for sigma_c
lo <- 0.15; hi <- 0.25
for (i in 1:40) {
  mid <- (lo+hi)/2
  if (length(chronic_equilibria(mid)) >= 2) lo <- mid else hi <- mid
}
cat(sprintf("sigma_c (numerically located) ~= %.5f\n", (lo+hi)/2))
```

## Block 3 — Bistability

```{r bistability, fig.width=8, fig.height=6.5}
jacobian_ih <- function(x, y, sigma) {
  dPdx <- nu1*(1-(2*x+y)/K) - mu1 - beta*y
  dPdy <- -nu1*x/K - beta*x
  dQdx <- sigma*beta*y - nu2*y/K
  dQdy <- sigma*beta*x + nu2*(1-(x+2*y)/K) - mu2
  matrix(c(dPdx, dQdx, dPdy, dQdy), nrow=2)
}
is_stable_ih <- function(x, y, sigma) {
  eig <- eigen(jacobian_ih(x,y,sigma), only.values=TRUE)$values
  list(stable = all(Re(eig) < 0), eig = eig)
}

sigma_test <- 0.1
coeffs_f1 <- c(-nu1/K, (nu1-mu1), lam)
roots_x0 <- polyroot(rev(coeffs_f1))
x0 <- max(Re(roots_x0)[Re(roots_x0) > 0])
R0 <- (1/mu2) * (sigma_test*beta*x0 + nu2*(1 - x0/K))
cat(sprintf("Testing sigma=%.2f, R0=%.4f (< 1)\n", sigma_test, R0))

res0 <- is_stable_ih(x0, 0, sigma_test)
cat(sprintf("P0=(%.2f,0): stable=%s\n", x0, res0$stable))

for (eq in chronic_equilibria(sigma_test)) {
  res <- is_stable_ih(eq[1], eq[2], sigma_test)
  cat(sprintf("Chronic equilibrium (%.2f,%.2f): stable=%s\n", eq[1], eq[2], res$stable))
}

starts <- list(c(830,2), c(700,50), c(500,150), c(300,300), c(100,500), c(50,650), c(18,610), c(200,20))
traj_list <- list()
for (i in seq_along(starts)) {
  pt <- starts[[i]]
  out <- as.data.frame(ode(y = c(x=pt[1], y=pt[2]), times = seq(0,3000,length.out=2000),
                             func = inhost_model, parms = c(sigma=sigma_test), method = "lsoda"))
  out$traj <- i
  traj_list[[i]] <- out
  cat(sprintf("start=(%g,%g) -> final=(%.2f,%.2f)\n", pt[1], pt[2], tail(out$x,1), tail(out$y,1)))
}
traj_df <- bind_rows(traj_list)

ggplot(traj_df, aes(x, y, group = traj)) +
  geom_path(alpha = 0.8) +
  annotate("point", x = x0, y = 0, color = "#3c763d", shape = 8, size = 5) +
  labs(title = sprintf("Bistability at sigma=%.1f (R0=%.3f < 1)", sigma_test, R0)) +
  theme_minimal()

ggsave("../../chapter_05/images/backward_bifurcation_bistability_R.png", width = 8, height = 6.5, dpi = 130)
```

## Block 4 — Dulac's criterion (numerical sign check, mirroring the Python symbolic proof)

```{r dulac}
# R's symbolic tools are less convenient here; the Python notebook gives the
# full symbolic derivation. We confirm the same conclusion numerically: the
# Dulac expression -lambda/(x^2*y) - nu1/(K*y) - nu2/(K*x) is negative
# everywhere it's evaluated.
dulac_expr <- function(x, y) -lam/(x^2*y) - nu1/(K*y) - nu2/(K*x)
test_pts <- expand.grid(x = seq(10, K, length.out=20), y = seq(10, K, length.out=20))
vals <- mapply(dulac_expr, test_pts$x, test_pts$y)
cat(sprintf("Dulac expression: max value on grid = %.6f (should be < 0)\n", max(vals)))
cat(sprintf("All negative? %s\n", all(vals < 0)))
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_05/section_5_2.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_05/exercises/section_5_2_exercises.md`; solutions:
`../../chapter_05/solutions/section_5_2_solutions.md`.*

---

# Chapter 5 Complete — and the Book Is Complete

This closes Chapter 5, and with it, the entire textbook.
