---
title: "Chapter 1, Section 1.2 — Deterministic Epidemic Models: Compartmental Approach"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 6–8. Companion to `../../python/chapter_01/section_1_2.ipynb` —
content mirrors that notebook block by block; see it for the full prose (rewrite, ELI10, key
ideas, symbol table, intuition) which is identical here. This file focuses on reproducing the code.

---

## Block 1 — Recreating the transfer diagram (Figure 1.1)

```{r transfer-diagram, fig.width=9, fig.height=3.6}
boxes <- tibble::tibble(label = c("S", "I", "R"), x = c(2, 5, 8), y = 1.8)

edges <- tibble::tribble(
  ~x, ~y, ~xend, ~yend, ~label, ~curve,
  2.6, 1.8, 4.4, 1.8, "new infections", 0,
  5.6, 1.8, 7.4, 1.8, "recovery", 0,
  7.4, 2.3, 2.6, 2.3, "loss of immunity", -0.3
)

ggplot() +
  geom_rect(data = boxes, aes(xmin = x - 0.6, xmax = x + 0.6, ymin = y - 0.6, ymax = y + 0.6),
            fill = "#eef2f7", color = "#31708f", linewidth = 1) +
  geom_text(data = boxes, aes(x = x, y = y, label = label), size = 6, fontface = "bold") +
  geom_curve(data = edges, aes(x = x, y = y, xend = xend, yend = yend),
             curvature = edges$curve, arrow = arrow(length = unit(0.2, "cm")), color = "black") +
  geom_text(data = edges, aes(x = (x + xend) / 2, y = pmax(y, yend) + 0.35, label = label), size = 3) +
  coord_cartesian(xlim = c(0, 10), ylim = c(0, 4)) +
  theme_void() +
  labs(title = "Figure 1.1 (recreated): Transfer diagram for an SIR compartment model")

ggsave("../../chapter_01/images/transfer_diagram_1_1_R.png", width = 9, height = 3.6, dpi = 130)
```

## Block 2 — Verifying the difference-quotient-to-derivative limit

```{r symbolic-check}
# R has no first-class symbolic engine as convenient as sympy for this; we
# verify the SAME claim purely numerically instead (equivalent rigor for our
# purposes -- see the Python notebook for the symbolic sympy::limit() version).
S_func <- function(tt) 1000 * exp(-0.05 * tt) + 50 * cos(tt)
Sprime_true <- function(tt) -50 * exp(-0.05 * tt) - 50 * sin(tt)

t0 <- 5.0
true_deriv <- Sprime_true(t0)

dts <- c(1.0, 0.1, 0.01, 0.001, 0.0001)
for (dt in dts) {
  dq <- (S_func(t0 + dt) - S_func(t0)) / dt
  cat(sprintf("dt=%8.4f | difference quotient = %10.6f | error = %.8f\n",
              dt, dq, abs(dq - true_deriv)))
}
cat(sprintf("\nTrue derivative S'(%.0f) = %.6f\n", t0, true_deriv))
```

```{r convergence-plot, fig.width=11, fig.height=4}
t_grid <- seq(0, 20, length.out = 400)
p1 <- ggplot(tibble(t = t_grid, S = S_func(t_grid)), aes(t, S)) +
  geom_line(color = "#31708f") +
  labs(title = "A concrete example S(t)") +
  theme_minimal()

dt_seq <- 10^seq(0, -4, length.out = 40)
errs <- sapply(dt_seq, function(d) abs((S_func(t0 + d) - S_func(t0)) / d - true_deriv))
p2 <- ggplot(tibble(dt = dt_seq, err = errs), aes(dt, err)) +
  geom_point(color = "#a94442") + geom_line(color = "#a94442") +
  scale_x_log10() + scale_y_log10() +
  labs(title = TeX("Convergence as $\\Delta t \\to 0$"), x = TeX("$\\Delta t$"),
       y = "| difference quotient - true derivative |") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
ggsave("../../chapter_01/images/difference_quotient_convergence_R.png",
       plot = gridExtra::arrangeGrob(p1, p2, ncol = 2), width = 11, height = 4, dpi = 130)
```

**Reading the output:** as in the Python notebook, the difference quotient converges to the true
derivative as $\Delta t$ shrinks — this is the numerical face of the Step 3 limiting argument in
the main derivation (see the Python notebook for the full step-by-step writeup, identical here).

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_01/section_1_2.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_01/exercises/section_1_2_exercises.md`; solutions:
`../../chapter_01/solutions/section_1_2_solutions.md`.*
