---
title: "Chapter 1, Section 1.4 — Important Concepts in Compartmental Epidemic Models"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
library(deSolve)
```

**Source:** Li, M.Y. (2018), pp. 11–29. Companion to `../../python/chapter_01/section_1_4.ipynb` —
full prose (rewrite, ELI10, key ideas, derivations, symbol tables, intuition) is identical to that
notebook; this file reproduces the four blocks' code in R.

---

## Block 1 (§1.4.1) — Exponential residence time: mean and memorylessness

```{r exp-distribution-checks}
# R lacks a convenient symbolic engine for the general derivation (see the Python
# notebook's sympy version for the fully symbolic proof); here we verify the
# SAME three claims numerically/via base R calculus utilities.
r_rate <- 0.2

f <- function(t, r) r * exp(-r * t)

# (1) integrates to 1
total_prob <- integrate(f, lower = 0, upper = Inf, r = r_rate)$value
cat("Integral of f(t):", total_prob, "\n")

# (2) mean = 1/r
mean_X <- integrate(function(t) t * f(t, r_rate), lower = 0, upper = Inf)$value
cat(sprintf("E[X] = %.6f  (should equal 1/r = %.6f)\n", mean_X, 1 / r_rate))

# (3) memoryless property, checked numerically at a few (s, t) pairs
G <- function(t, r) exp(-r * t)
for (s in c(1, 5, 10)) {
  for (t in c(1, 3, 7)) {
    lhs <- G(s + t, r_rate) / G(s, r_rate)
    rhs <- G(t, r_rate)
    cat(sprintf("s=%2d, t=%2d: G(s+t)/G(s) - G(t) = %.10f\n", s, t, lhs - rhs))
  }
}
```

## Block 1 continued — ODE vs. DDE residence-time comparison

```{r ode-vs-dde, fig.width=9, fig.height=4.5}
lambda_ <- 0.0005
gamma_ <- 0.2
S0 <- 550; I0 <- 5
t_max <- 120

# (A) ODE version, via deSolve
kmk <- function(t, y, parms) {
  S <- y["S"]; I <- y["I"]
  dS <- -parms["lambda_"] * I * S
  dI <- parms["lambda_"] * I * S - parms["gamma_"] * I
  list(c(dS, dI))
}
t_eval <- seq(0, t_max, length.out = 600)
sol_ode <- ode(y = c(S = S0, I = I0), times = t_eval, func = kmk,
                parms = c(lambda_ = lambda_, gamma_ = gamma_), method = "lsoda")
sol_ode <- as.data.frame(sol_ode)

# (B) DDE version via a windowed-sum (rectangle-rule integral) formulation --
# mirrors the Python notebook's stable implementation exactly (a naive
# difference-equation version of this DDE is numerically unstable; see the
# Python notebook's comments for why).
omega <- 1 / gamma_
dt <- 0.02
n_steps <- as.integer(t_max / dt)
history_len <- as.integer(round(omega / dt))

S_dde <- numeric(n_steps + 1); I_dde <- numeric(n_steps + 1)
S_dde[1] <- S0; I_dde[1] <- I0
queue <- numeric(0)

for (k in seq_len(n_steps)) {
  Sk <- S_dde[k]; Ik <- I_dde[k]
  inc_now <- lambda_ * Ik * Sk
  packet <- inc_now * dt
  queue <- c(queue, packet)

  outflow <- 0
  while (length(queue) > history_len) {
    outflow <- outflow + queue[1]
    queue <- queue[-1]
  }

  S_dde[k + 1] <- max(Sk - inc_now * dt, 0)
  I_dde[k + 1] <- max(Ik + packet - outflow, 0)
}
t_dde <- seq(0, t_max, length.out = n_steps + 1)

plot_df <- bind_rows(
  tibble(t = sol_ode$time, I = sol_ode$I, type = "exponential residence time (ODE)"),
  tibble(t = t_dde, I = I_dde, type = sprintf("fixed residence time omega=%.0f (DDE)", omega))
)

ggplot(plot_df, aes(x = t, y = I, color = type, linetype = type)) +
  geom_line(linewidth = 1) +
  labs(title = "Same mean infectious period, two different residence-time distributions",
       x = "time", y = "I(t)") +
  theme_minimal()

ggsave("../../chapter_01/images/ode_vs_dde_residence_time_R.png", width = 9, height = 4.5, dpi = 130)

cat(sprintf("Peak I(t), exponential (ODE): %.2f\n", max(sol_ode$I)))
cat(sprintf("Peak I(t), fixed omega (DDE): %.2f\n", max(I_dde)))
```

## Block 2 (§1.4.2) — Comparing incidence functional forms

```{r incidence-forms, fig.width=8.5, fig.height=5}
I_grid <- seq(0, 500, length.out = 300)
S_fixed <- 800
beta <- 0.0005
lam <- 0.4
N_fixed <- 1500
K <- 100

incidence_df <- tibble(
  I = I_grid,
  mass_action = beta * I_grid * S_fixed,
  saturating = beta * I_grid * S_fixed / (K + I_grid),
  standard = lam * I_grid * S_fixed / N_fixed
) %>%
  pivot_longer(-I, names_to = "form", values_to = "rate")

ggplot(incidence_df, aes(x = I, y = rate, color = form)) +
  geom_line(linewidth = 1) +
  labs(title = sprintf("Comparing incidence functional forms (S=%d fixed)", S_fixed),
       x = "I (number infectious)", y = "incidence rate") +
  theme_minimal()

ggsave("../../chapter_01/images/incidence_forms_comparison_R.png", width = 8.5, height = 5, dpi = 130)

cat("As I grows large, saturating incidence approaches beta*S =", beta * S_fixed, "\n")
```

## Block 3 (§1.4.3) — Demography and logistic growth

```{r demography, fig.width=14, fig.height=4}
sir_demography <- function(t, y, parms) {
  S <- y["S"]; I <- y["I"]; R <- y["R"]
  N <- S + I + R
  b <- parms["b"]; d1 <- parms["d1"]; d2 <- parms["d2"]; d3 <- parms["d3"]
  lam <- parms["lam"]; gamma_ <- parms["gamma_"]
  dS <- b*N - lam*I*S - d1*S
  dI <- lam*I*S - (gamma_ + d2)*I
  dR <- gamma_*I - d3*R
  list(c(dS, dI, dR))
}

t_eval2 <- seq(0, 200, length.out = 500)
cases <- list(
  list(label = "b > d (growing)", b = 0.02, d = 0.01),
  list(label = "b < d (shrinking)", b = 0.01, d = 0.02),
  list(label = "b = d (constant N)", b = 0.015, d = 0.015)
)

demo_results <- purrr::map_dfr(cases, function(cs) {
  parms <- c(b = cs$b, d1 = cs$d, d2 = cs$d, d3 = cs$d, lam = lambda_, gamma_ = gamma_)
  out <- as.data.frame(ode(y = c(S = S0, I = I0, R = 0), times = t_eval2,
                             func = sir_demography, parms = parms, method = "lsoda"))
  out$N <- out$S + out$I + out$R
  out$label <- cs$label
  out
})

ggplot(demo_results, aes(x = time)) +
  geom_line(aes(y = N, color = "N(t)")) +
  geom_line(aes(y = I, color = "I(t)")) +
  facet_wrap(~label, scales = "free_y") +
  labs(title = "Demographic SIR model: equal death rates -> exponential N(t)",
       x = "time", y = "count", color = "") +
  theme_minimal()

ggsave("../../chapter_01/images/demography_sir_cases_R.png", width = 14, height = 4, dpi = 130)

# Verification against closed-form N0*exp((b-d)t)
grow_case <- demo_results %>% filter(label == "b > d (growing)")
N_closed_form <- (S0 + I0) * exp((0.02 - 0.01) * grow_case$time)
max_rel_error <- max(abs(grow_case$N - N_closed_form) / N_closed_form)
cat(sprintf("Max relative error vs closed-form N0*exp((b-d)t): %.2e\n", max_rel_error))
```

```{r logistic-growth, fig.width=8, fig.height=4.5}
sir_logistic <- function(t, y, parms) {
  S <- y["S"]; I <- y["I"]; R <- y["R"]
  N <- S + I + R
  b <- parms["b"]; d <- parms["d"]; K <- parms["K"]; lam <- parms["lam"]; gamma_ <- parms["gamma_"]
  dS <- b*N - (N^2)/K - lam*I*S - d*S
  dI <- lam*I*S - (gamma_ + d)*I
  dR <- gamma_*I - d*R
  list(c(dS, dI, dR))
}

b <- 0.05; d <- 0.01; K <- 2000
t_eval3 <- seq(0, 300, length.out = 600)
sol_log <- as.data.frame(ode(y = c(S = S0, I = I0, R = 0), times = t_eval3,
                               func = sir_logistic, parms = c(b = b, d = d, K = K, lam = lambda_, gamma_ = gamma_),
                               method = "lsoda"))
sol_log$N <- sol_log$S + sol_log$I + sol_log$R
carrying_capacity <- (b - d) * K

ggplot(sol_log, aes(x = time, y = N)) +
  geom_line(color = "#31708f") +
  geom_hline(yintercept = carrying_capacity, linetype = "dashed", color = "grey40") +
  labs(title = "Logistic population growth approaches (b-d)K",
       x = "time", y = "N(t)") +
  theme_minimal()

ggsave("../../chapter_01/images/logistic_growth_carrying_capacity_R.png", width = 8, height = 4.5, dpi = 130)

cat(sprintf("Final simulated N(t_end) = %.1f, predicted carrying capacity (b-d)K = %.1f\n",
            tail(sol_log$N, 1), carrying_capacity))
```

## Block 4 (§1.4.4) — SEIR preview

```{r seir-preview, fig.width=9, fig.height=4.5}
seir <- function(t, y, parms) {
  S <- y["S"]; E <- y["E"]; I <- y["I"]; R <- y["R"]
  lam <- parms["lam"]; kappa <- parms["kappa"]; gamma_ <- parms["gamma_"]
  dS <- -lam*I*S
  dE <- lam*I*S - kappa*E
  dI <- kappa*E - gamma_*I
  dR <- gamma_*I
  list(c(dS, dE, dI, dR))
}

kappa <- 0.15
t_eval4 <- seq(0, 150, length.out = 600)
sol_seir <- as.data.frame(ode(y = c(S = S0, E = 0, I = I0, R = 0), times = t_eval4,
                                func = seir, parms = c(lam = lambda_, kappa = kappa, gamma_ = gamma_),
                                method = "lsoda"))

seir_long <- sol_seir %>% pivot_longer(cols = c("S", "E", "I", "R"), names_to = "compartment", values_to = "count")

ggplot(seir_long, aes(x = time, y = count, color = compartment)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c(S = "#31708f", E = "#dca61a", I = "#a94442", R = "#3c763d")) +
  labs(title = sprintf("SEIR preview: mean latent period = %.1f", 1/kappa),
       x = "time", y = "count") +
  theme_minimal()

ggsave("../../chapter_01/images/seir_preview_R.png", width = 9, height = 4.5, dpi = 130)

N_vals <- sol_seir$S + sol_seir$E + sol_seir$I + sol_seir$R
cat(sprintf("N(0)=%.4f, N(t_end)=%.4f, max deviation=%.2e\n", N_vals[1], tail(N_vals,1), max(abs(N_vals - N_vals[1]))))
peak_E_time <- sol_seir$time[which.max(sol_seir$E)]
peak_I_time <- sol_seir$time[which.max(sol_seir$I)]
cat(sprintf("Peak E(t) at t=%.1f, peak I(t) at t=%.1f\n", peak_E_time, peak_I_time))
```

## Summary / Memory Aids / Connections

Identical to `../../python/chapter_01/section_1_4.ipynb` — see that notebook for the full text.

*Practice problems: `../../chapter_01/exercises/section_1_4_exercises.md`; solutions:
`../../chapter_01/solutions/section_1_4_solutions.md`.*
