---
title: "Chapter 1, Section 1.1 — Mathematical Modeling of Infectious Diseases: Issues and Approaches"
author: "Mathematical Modeling Study"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
library(ggplot2)
library(dplyr)
library(tidyr)
library(pracma)
library(latex2exp)
# deSolve is loaded for consistency with later sections; Sec. 1.1 has no ODEs yet.
library(deSolve)
```

**Source:** Li, M.Y. (2018), *An Introduction to Mathematical Modeling of Infectious Diseases*,
Springer, pp. 1–6. See `../python/section_1_1.ipynb` for the companion Python notebook — this
file mirrors it exactly, block by block.

Section 1.1 is entirely conceptual (no formulas yet), so the code chunks below build small
illustrative tools rather than solving equations. Real ODEs start in `section_1_2.Rmd`.

---

## Block 1 — What makes a disease "infectious," and the vocabulary of spread

### Rewrite (plain English)

A disease counts as *infectious* if whatever causes it can move from one host to another —
through touch, air, water/food, a disease vector, or mother-to-child. A model of an infectious
disease describes how it spreads through a population: infectious people mix with susceptible
people and the disease passes between them. Depending on scale, speed, persistence, and
geography, we call the resulting spread an **outbreak**, **epidemic**, **endemic** disease, or
**pandemic** (the book's example: the 1918 Spanish flu, which killed over 50 million people
across every continent).

### Explain Like I'm 10

Same rumor-at-school story as the Python notebook: an outbreak is more kids than usual hearing it
suddenly; an epidemic is it ripping through the whole school fast; endemic is it never quite going
away because new kids keep arriving; pandemic is it jumping to other schools, then other
countries.

### Key Ideas

Same table as the Python notebook: mode of transmission, susceptible/infectious roles,
asymptomatic infection, the outbreak/epidemic/endemic/pandemic distinction, and acquired
immunity — see `../python/section_1_1.ipynb` Block 1 for the full table and why each matters
later in the book.

### Visual Understanding — classification logic + flow diagram

```{r classify-function}
# Classify a disease event using ONLY the verbal definitions in Sec. 1.1.
# (Mirrors classify_disease_event() in the Python notebook exactly.)
classify_disease_event <- function(above_baseline, fast_widespread,
                                    persists_long_term, crosses_continents) {
  if (!above_baseline) return("no event")
  if (crosses_continents) return("pandemic")
  if (persists_long_term) return("endemic")
  if (fast_widespread) return("epidemic")
  return("outbreak")
}

# Sanity-check against the book's own example: 1918 Spanish flu
spanish_flu <- classify_disease_event(
  above_baseline = TRUE, fast_widespread = TRUE,
  persists_long_term = FALSE, crosses_continents = TRUE
)
cat("1918 Spanish flu classifies as:", spanish_flu, "\n")

local_cluster <- classify_disease_event(
  above_baseline = TRUE, fast_widespread = FALSE,
  persists_long_term = FALSE, crosses_continents = FALSE
)
cat("Small local case cluster classifies as:", local_cluster, "\n")

persistent_regional <- classify_disease_event(
  above_baseline = TRUE, fast_widespread = FALSE,
  persists_long_term = TRUE, crosses_continents = FALSE
)
cat("Persistent regional transmission classifies as:", persistent_regional, "\n")
```

```{r verification-block1}
# Verification: same five assertions as the Python notebook
stopifnot(classify_disease_event(FALSE, FALSE, FALSE, FALSE) == "no event")
stopifnot(classify_disease_event(TRUE, TRUE, FALSE, TRUE) == "pandemic")
stopifnot(classify_disease_event(TRUE, FALSE, TRUE, FALSE) == "endemic")
stopifnot(classify_disease_event(TRUE, TRUE, FALSE, FALSE) == "epidemic")
stopifnot(classify_disease_event(TRUE, FALSE, FALSE, FALSE) == "outbreak")
cat("All 5 classification rules verified against Sec. 1.1 definitions.\n")
```

```{r flow-diagram, fig.width=9, fig.height=5}
# Flow diagram of the decision logic, built with base ggplot2 geoms
# (rectangles + text + segments), mirroring the matplotlib version in Python.
boxes <- tibble::tribble(
  ~key,        ~x,   ~y,   ~label,                                   ~kind,
  "start",     0.08, 0.5,  "Case count rises\nabove baseline?",      "decision",
  "no",        0.08, 0.08, "No event",                               "no",
  "cont",      0.35, 0.85, "Crosses\ncontinents?",                   "decision",
  "pandemic",  0.62, 0.85, "PANDEMIC",                                "yes",
  "persist",   0.35, 0.5,  "Keeps circulating\n(new susceptibles /\neasy reinfection)?", "decision",
  "endemic",   0.62, 0.5,  "ENDEMIC",                                 "yes",
  "fast",      0.35, 0.15, "Spreads fast to\nmany people?",          "decision",
  "epidemic",  0.62, 0.15, "EPIDEMIC",                                "yes",
  "outbreak",  0.88, 0.15, "OUTBREAK\n(neither fast\nnor global)",   "yes"
)

edges <- tibble::tribble(
  ~x, ~y, ~xend, ~yend,
  0.08, 0.5,  0.08, 0.08,   # start -> no
  0.08, 0.5,  0.35, 0.85,   # start -> cont
  0.35, 0.85, 0.62, 0.85,   # cont -> pandemic
  0.35, 0.85, 0.35, 0.5,    # cont -> persist
  0.35, 0.5,  0.62, 0.5,    # persist -> endemic
  0.35, 0.5,  0.35, 0.15,   # persist -> fast
  0.35, 0.15, 0.62, 0.15,   # fast -> epidemic
  0.35, 0.15, 0.88, 0.15    # fast -> outbreak
)

fill_map <- c(decision = "#eef2f7", no = "#f2dede", yes = "#dff0d8")

ggplot() +
  geom_segment(data = edges, aes(x = x, y = y, xend = xend, yend = yend),
               arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  geom_label(data = boxes, aes(x = x, y = y, label = label, fill = kind),
             color = "black", label.size = 0.3, size = 3) +
  scale_fill_manual(values = fill_map, guide = "none") +
  theme_void() +
  labs(title = "Classifying a disease event (Sec. 1.1 vocabulary)") +
  theme(plot.title = element_text(hjust = 0.5, size = 13))

ggsave("../../chapter_01/images/disease_vocabulary_flow_R.png", width = 9, height = 5, dpi = 130)
```

---

## Block 2 — The questions public health authorities actually ask

Rewrite / ELI10 / Key Ideas / Intuition: identical content to the Python notebook Block 2 — see
`../python/section_1_1.ipynb` for the full text (severity, duration, intervention effectiveness,
stockpiling, containment; and the four reasons traditional methods fall short: scale, ethics,
data completeness, single-sample problem).

```{r scale-mismatch}
# Same "scale mismatch" illustration as the Python notebook: potential contact
# pairs grow quadratically with population size, not linearly.
population_sizes <- c(10, 100, 1000, 10000, 100000)
potential_contact_pairs <- population_sizes * (population_sizes - 1) / 2

scale_df <- tibble::tibble(
  population_size = population_sizes,
  potential_contact_pairs = potential_contact_pairs,
  pairs_per_person = potential_contact_pairs / population_sizes
)
scale_df
```

**Reading the table:** identical takeaway to the Python version — a city of 100,000 has roughly
5 billion potential contact pairs, which is why lab-scale experiments cannot capture population-
scale transmission dynamics.

---

## Block 3 — What modeling provides, and the six-stage modeling process

Rewrite / ELI10 / Key Ideas / Intuition: identical content to the Python notebook Block 3 (the
paper-airplane analogy and the six-stage table). See `../python/section_1_1.ipynb`.

```{r six-stage-loop, fig.width=7, fig.height=7}
stages <- c(
  "1. Make\nassumptions",
  "2. Build model\n(diagram -> equations)",
  "3. Mathematical\nanalysis",
  "4. Interpret\nfindings",
  "5. Collect data\n& validate",
  "6. Improve\nassumptions"
)

n <- length(stages)
angles <- seq(pi / 2, pi / 2 - 2 * pi, length.out = n + 1)[1:n]  # start at top, go clockwise
radius <- 1
loop_df <- tibble::tibble(
  stage = stages,
  x = radius * cos(angles),
  y = radius * sin(angles)
)

# Build arrows stage i -> stage i+1 (looping 6 -> 1), shrunk so they don't
# overlap the label boxes.
shrink <- 0.30
edge_df <- purrr::map_dfr(seq_len(n), function(i) {
  j <- if (i == n) 1 else i + 1
  x0 <- loop_df$x[i]; y0 <- loop_df$y[i]
  x1 <- loop_df$x[j]; y1 <- loop_df$y[j]
  dx <- x1 - x0; dy <- y1 - y0
  dist <- sqrt(dx^2 + dy^2)
  tibble::tibble(
    x = x0 + dx * shrink / dist, y = y0 + dy * shrink / dist,
    xend = x1 - dx * shrink / dist, yend = y1 - dy * shrink / dist
  )
})

ggplot() +
  geom_curve(data = edge_df, aes(x = x, y = y, xend = xend, yend = yend),
             curvature = 0.15, arrow = arrow(length = unit(0.2, "cm")), color = "grey40") +
  geom_label(data = loop_df, aes(x = x, y = y, label = stage),
             fill = "#eef2f7", color = "#31708f", size = 3.2) +
  coord_equal(xlim = c(-1.6, 1.6), ylim = c(-1.6, 1.6)) +
  theme_void() +
  labs(title = "The six-stage mathematical modeling process (Sec. 1.1)") +
  theme(plot.title = element_text(hjust = 0.5, size = 13))

ggsave("../../chapter_01/images/six_stage_modeling_loop_R.png", width = 7, height = 7, dpi = 130)
```

---

## Block 4 — A model is an approximation; limitations; the realism/tractability trade-off

Rewrite / ELI10 / Key Ideas / Intuition: identical content to the Python notebook Block 4 (the
map-vs-city analogy, the three named limitations, and the complexity/tractability trade-off).

```{r realism-tradeoff, fig.width=11, fig.height=4.2}
set.seed(42)

x <- seq(0, 10, length.out = 8)
y_true <- 2.0 + 0.5 * x
y_obs <- y_true + rnorm(length(x), sd = 1.0)

x_grid <- seq(0, 10, length.out = 100)
n_boot <- 200
simple_curves <- matrix(NA, nrow = n_boot, ncol = length(x_grid))
complex_curves <- matrix(NA, nrow = n_boot, ncol = length(x_grid))

for (b in seq_len(n_boot)) {
  idx <- sample(seq_along(x), length(x), replace = TRUE)
  xb <- x[idx]; yb <- y_obs[idx]
  simple_fit <- lm(yb ~ xb)                                  # 2 parameters
  complex_fit <- lm(yb ~ poly(xb, 6, raw = TRUE))             # 7 parameters
  simple_curves[b, ] <- predict(simple_fit, newdata = data.frame(xb = x_grid))
  complex_curves[b, ] <- predict(complex_fit, newdata = data.frame(xb = x_grid))
}

simple_spread <- mean(apply(simple_curves, 2, sd))
complex_spread <- mean(apply(complex_curves, 2, sd))

cat(sprintf("Average prediction uncertainty, SIMPLE model (2 params):  %.3f\n", simple_spread))
cat(sprintf("Average prediction uncertainty, COMPLEX model (7 params): %.3f\n", complex_spread))
cat(sprintf("\nThe more 'realistic' (higher-degree) model is %.1fx more uncertain, from the SAME 8 data points.\n",
            complex_spread / simple_spread))

# Plot
plot_df <- bind_rows(
  as_tibble(t(simple_curves[seq(1, n_boot, by = 4), ])) %>%
    mutate(x = x_grid) %>%
    pivot_longer(-x, names_to = "rep", values_to = "y") %>%
    mutate(model = "Simple (2 params)"),
  as_tibble(t(complex_curves[seq(1, n_boot, by = 4), ])) %>%
    mutate(x = x_grid) %>%
    pivot_longer(-x, names_to = "rep", values_to = "y") %>%
    mutate(model = "Complex (7 params)")
)
obs_df <- tibble(x = x, y = y_obs)

ggplot(plot_df, aes(x = x, y = y, group = rep)) +
  geom_line(alpha = 0.08, color = ifelse(plot_df$model == "Simple (2 params)", "#31708f", "#a94442")) +
  geom_point(data = obs_df, aes(x = x, y = y, group = NULL), color = "black", size = 2) +
  facet_wrap(~model) +
  coord_cartesian(ylim = c(-5, 12)) +
  labs(title = "Realism vs. tractability: same data, two model complexities",
       x = "x", y = "y") +
  theme_minimal()

ggsave("../../chapter_01/images/realism_tradeoff_R.png", width = 11, height = 4.2, dpi = 130)
```

---

## Block 5 — Three modeling paradigms: statistical, deterministic, stochastic

Rewrite / ELI10 / Key Ideas / Intuition: identical content to the Python notebook Block 5 (the
three-friends-predicting-a-rumor analogy and the comparison table).

```{r det-vs-stoch, fig.width=11, fig.height=4.2}
set.seed(7)

simulate_birth_process <- function(N0, rate, t_max) {
  # Continuous-time linear birth process via the Gillespie algorithm.
  # Mirrors simulate_birth_process() in the Python notebook exactly.
  t <- 0
  N <- N0
  times <- c(t)
  pops <- c(N)
  while (t < t_max) {
    total_rate <- rate * N
    if (total_rate <= 0) break
    dt <- rexp(1, rate = total_rate)
    t <- t + dt
    if (t > t_max) break
    N <- N + 1
    times <- c(times, t)
    pops <- c(pops, N)
  }
  times <- c(times, t_max)
  pops <- c(pops, N)
  list(times = times, pops = pops)
}

rate <- 0.3
t_max <- 15
n_realizations <- 25
t_grid <- seq(0, t_max, length.out = 200)

make_panel_df <- function(N0, label) {
  deterministic <- tibble(t = t_grid, N = N0 * exp(rate * t_grid), label = label)
  sims <- purrr::map_dfr(seq_len(n_realizations), function(i) {
    sim <- simulate_birth_process(N0, rate, t_max)
    tibble(t = sim$times, N = sim$pops, rep = i, label = label)
  })
  list(det = deterministic, sims = sims)
}

small <- make_panel_df(5, "SMALL population (N0=5)")
large <- make_panel_df(500, "LARGE population (N0=500)")

sims_all <- bind_rows(small$sims, large$sims)
det_all <- bind_rows(small$det, large$det)

ggplot() +
  geom_step(data = sims_all, aes(x = t, y = N, group = rep),
            color = "#a94442", alpha = 0.25) +
  geom_line(data = det_all, aes(x = t, y = N), color = "#31708f", linewidth = 1.2) +
  facet_wrap(~label, scales = "free_y") +
  labs(title = "Why deterministic models need large populations",
       x = "time", y = "population size") +
  theme_minimal()

ggsave("../../chapter_01/images/deterministic_vs_stochastic_R.png", width = 11, height = 4.2, dpi = 130)

cat("Small-population runs spread widely around the deterministic curve;\n")
cat("large-population runs hug it closely -- exactly the trade-off the text describes.\n")
```

```{r verification-block5}
# Numerical verification: mean tracks deterministic curve; relative spread shrinks with N0.
for (N0 in c(5, 50, 500)) {
  finals <- replicate(300, {
    sim <- simulate_birth_process(N0, rate, t_max)
    tail(sim$pops, 1)
  })
  deterministic_final <- N0 * exp(rate * t_max)
  mean_ratio <- mean(finals) / deterministic_final
  coeff_of_variation <- sd(finals) / mean(finals)
  cat(sprintf("N0=%4d:  mean(stochastic)/deterministic = %5.3f   coefficient of variation = %5.3f\n",
              N0, mean_ratio, coeff_of_variation))
}
cat("\nAs N0 grows, the mean ratio -> 1 and the coefficient of variation shrinks.\n")
```

---

## Summary

Identical to the Python notebook's Summary section:

**One sentence:** Infectious disease modeling exists because traditional experimental and
statistical methods can't safely or completely answer public health's most urgent questions about
an epidemic, and mathematical models — built through a six-stage assumption→equations→analysis→
validation loop — fill that gap, at the cost of always being a simplified, imperfect approximation
of reality.

See `../python/section_1_1.ipynb` for the full one-paragraph summary, bullet points, memory aids,
and connections — content is identical across both languages by design, so you can study from
whichever you're running.

*Practice problems for this section are in `../../chapter_01/exercises/section_1_1_exercises.md`, with full
worked solutions in `../../chapter_01/solutions/section_1_1_solutions.md`.*
